---
title: "Introduction To R"
author: "Yuan Huang"
date: "December 21, 2016"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

This is the extracted content of "Introduction to R". The PDF file is saved as "Rcourse.pdf" 

### str can show the object structure
```{r str}
s<-list(id="example",sequence="ACCTAGAT",alphabet=c("A","C","G","T"))
class(s)<-"GenericSeq"
str(s)
```

### for while and repeat
for(var in sequence) expr
while (condition) expr
repeat expr

use print() and cat() to print the object inside the loop
use break and next to jump out of the loop, or to the next iteration

### sort() and order()
sort() is used to sort a vector. order() returns the index of the ordered vector, and can be used to reorder matrix by one of its columns.
```{r sort and order}
x=c(19,25,31,15,12,43,82,22)
sort(x)

mat=matrix(c(12,3,7,9,10,11),ncol=2)
mat
order(mat[,1])
mat[order(mat[,1]),]
```



### both ls() and objects() list the objects in the current enviornment
```{r ls}
ls()
objects()
```

###library() shows all the library installed on the system
```{r library}
library()
```

### Search path and objects
using search() to show the search path and objects(pos=num) to show all the objects in the designated directory in the search path
```{r search and objects}
search()
objects(pos=3)
```

### attach
using attach to add a database to the search path. To make objects from the previous session of R available, pass attach() the location of the .RData file. You can also directly access the dataframe or list elements by attaching the dataframe and list name, then using the element name to access it.
```{r attach}
df=data.frame(matrix(1:100,nrow=20))
colnames(df)=c("first","second","third","fourth","fifth")
attach(df)
mean(first)
```

### nchar 
nchar returns the nubmer of characters for character data and the printed representation of numeric data types. length() returns the number of elements. For matrix, it returns the total number of the elements in the matrix, while for data frame it returns the number of columns. dim() is used for arrays(dataframe and matrix). For non array objects,it returns null
```{r nchar}
nchar("abcdefg")
length("abcdefg")
length(matrix(1:100,nrow=20))
```

###objects
For objects, pos defines which working database to search, and pat defines the pattern. Setting the all.names=TRUE will display obj names begin with a period. apropos() accept a regular expression, and returns all the object anywhere in your search path wich match the expression

###get
get provides an alternative way to search objects. Suppose an object named x in a database "rproject/.Rdata", which is the 2nd in the search path, we can use either 
attach("rproject/.Rdata")
search()
or
get("x",2)

### c() and list()
c() combines objects in the most general way. A matrix combines with a vector results in a vector. list() function preserves the identity of each of its elements. When c() function is applied to lists, it will return a list. To break down the list into its individual components, use the recursive=TRUE argument of c()
```{r c()}
c(matrix(1:4, ncol=2),1:3)
list(matrix(1:4,ncol=2),1:3)
c(list(matrix(1:4,ncol=2),1:3),list(1:5))
unlist(c(list(matrix(1:4,ncol=2),1:3),list(1:5)))
c(list(matrix(1:4,ncol=2),1:3),list(1:5),recursive=TRUE)
```

### subscripting by []
```{r subscripting}
x=1
xv=c(1:10)
xv
xv[1:5]
xv[-3]
xv[-1:-3]
xv[c(-1,-3)]
xv[]=1
xv
```

### dimnames(), rownames() and colnames() and drop argument
```{r dim_row_col_names}
x<-matrix(1:15,nrow=5)
dimnames(x)<-list(NULL,c("one","two","three"))
x[3,2]
x[,1]
x[3,]
x[3:5,c(1,3)]
x[-c(1,3,5),]
x[x[,1]>2,]
x[,c("one","three")]

x[,3] # only a vector is available
x[,3,drop=FALSE] # you get the matrix with the column name
```

### the use of [[ in list
since the [ ] operation keeps the mode of their parent. For lists, R will often have problems working directly with such objects
```{r [[}
mylist<-list(a=1:10,b=10:20,c=30:40)
mean(mylist[1])
mean(mylist[[1]])
mean(mylist$a)
mean(mylist[["a"]])
```

### subscript in dataframe
```{r dataframe subsript}
df=data.frame(matrix(1:100,nrow=20))
colnames(df)=c("first","second","third","fourth","fifth")

df[["first"]]
df[,"first"]
df$first
df["first"] # a data frame with one column
df[1,"first"]
df[1,]$first
df[1,]["first"]
```

### using matrix elements as index
For matrix a, its rows are interpret as row and column numbers for indexing
```{r matrix index}
a=matrix(c(1,1,1,2,2,3),ncol=2,byrow=TRUE)
a
x=matrix(0,nrow=4,ncol=3)
x[a]=c(10,20,30)
```
since a has three rows, each row refers to an index combination.

### functions for vectors
```{r vector}
vc <-c(1:10)
vc
unique(vc)
duplicated(c(vc,1:5))
rev(vc)
sort(vc)
append(vc,c(1:5))
sum(vc)
min(vc)
max(vc)
```

### missing value
```{r missing value}
values=c(12,NA,19,15,12,17,14,NA,19)
values[!is.na(values)]
vv=matrix(values,ncol=3,byrow=TRUE)
vv[!is.na(vv[,2]),,drop=FALSE]
```

###soruce() and scan()
source can read the r source code and execute the code. scan() can read the content of file to R object.

###cat(), sink() and capture.output() NEED EXAMPLES

###recycle of vectors
```{r recycle}
x=matrix(1:3,nrow=2,ncol=6)
x
```

### %in% and in used in for loop NEED TO BE TESTED

### identical() test if two objects are exactly the same. all.equal() tests with a tolerance

### which
```{r which}
vt<-c(TRUE,FALSE, TRUE, FALSE, TRUE, TRUE)
which(vt)
which(vt, arr.ind=FALSE)
```

###cut function can convert the vectors to the designated categorical variables with labels or ranges
```{r cut}
x=c(17,19,22,43,14,8,12,19,20,51,8,12,27,31,44)
cut(x,3)
cut(x,3,labels=c("low","medium","high"))

#set the breaking points
cut(x,breaks=quantile(x,c(0,.25,.50,.75,1)),labels=c("Q1","Q2","Q3","Q4"),include.lowest = TRUE)
```

### the rep function
```{r rep}
rep(1:3,3)
rep(1:3,c(5,2,1))
```

### table() function
table() gives the frequency table of data
```{r table function}
x=c(7,12,19,7,19,21,12,14,17,12,19,21)
t1<-table(x)
sum(table(x)*as.numeric(names(table(x))))

x1<-c(1,2,3,2,1,3,2,3,1)
x2<-c(2,1,2,3,1,3,2,2,3)
x3<-c(1,2,3,3,2,1,2,1,2)
table(x1,x2)
ftable(table(x1,x2,x3))
table(table(x1,x2,x3))  # among the 27 permutations, how many appeared in the data
```

### Date values
```{r date}
d1<-as.Date("20041202",format="%Y%m%d")
d2<-as.Date("12/5/03", format="%m/%d/%y")
d3<-as.Date("7-8-2001",format="%d-%m-%Y")

d1-d2
d1-d2 > 360
mean(c(d1,d2,d3))
d1>d2
```
### chron
chron obj accepts two arguments, one for date, and the other for time. When in m/d/y and h:m:s format, the formats= argument is not needed

```{r chron}
library(chron)
d1=chron('12/25/2004','0:0:0')

d2=chron(10,origin=c(12,31,2004))  #origin accepts three arguments for m,d,Y, and add the 1st argument(10) days
d3=chron('7-6-2005','6:00:00',format=c(dates='m-d-y',times='h:m:s'))
d3-d1
weekdays(d3)
```


### file operations
```{r files}
file.info("generic_Seq.R")
file.access("generic_Seq.R")
list.files()
#file.choose()  #interactive user inteface for choosing files

#file operations
file.create("new_test_file.txt")
file.create("file1")
file.copy("file1","file2")
file.rename("new_test_file.txt","rnew_file.txt")
#file.symlink()  need to find out
file.exists("rnew_file.txt")
file.append("file1","rnew_file.txt")
dir.create("new_dir")
file.remove("file2")
file.show("file1")
```
###system operations
```{r system ops}
system.time(124*234*345)  #returns the time for evaluating the expr
proc.time()  #returns the user, system and total time for the current R process and its children
Sys.info()
Sys.time() # returns the current time
```

### Need to figure out what is <<-

### matrix operations
```{r matrix}
m1<-matrix(1:12,nrow=3)
m2<-matrix(101:112,nrow=4)
m1%*%m2
t(m1)
rbind(m1,c(1,2,3,4))
#solve(m2,c(90,100,110,120))
#solve(t(m2)%*%m2)
diag(m2)
diag(3)
row(m2)
col(m2)
```

### Generating sequences
```{r seq}
seq(1,10,by=2)
seq(1,10,length=3)
cv<-c(1:6)
seq(along=cv)
```

### sampling
sample command can set up replace, size, and prob for each poplulation sample
```{r sampling}
sample(10)
sample(c("a","b","c","d","e"),size=10,replace=TRUE)
sample(c("a","b","c"),size=10,prob=c(.5,.25,.25),replace=TRUE)
```

### functional Programming
apply() - operate on selected dimensions of arrays
lapply(), sapply() - operate on each element of a vector or list
tapply() - operate on elements based on a grouping variable
mapply() - multivariate extension of sapply
```{r functional programming}
mt1=matrix(1:100,nrow=20,dimnames=list(NULL,c("first","second","third","fourth","fifth")))
apply(mt1,1,mean)  #row mean
apply(mt1,2, mean) #column mean
```

#### tapply
tapply and aggregate both can output the function results for dataframe using group by, but tapply outputs tables/matrix, while aggregate outputs dataframes. In addition, the grouping column for aggregate must be a list.
```{r tapply}
interest.rate<-sample(rnorm(10),size=10,replace=TRUE)
state<-sample(c("AZ","TX","MA","WA","NV"),size=10, replace=TRUE)
name<-c("Bank of America","Wells Fargo","Nation Bank","Credit Union", "Citi","1st Bank","Mountain Valley", 
        "US Bank","Chase","West Bank")
banks<-data.frame(name,state,interest.rate)
tapply(banks$interest.rate,banks$state,max, na.rm=TRUE)
with(banks, tapply(interest.rate,state,max,na.rm=TRUE))
with(banks,aggregate(interest.rate,list(state),max,na.rm=TRUE)) #remeber to include state column in a list

b1<-banks[,c("state","interest.rate")]

testdata = data.frame(one=c(1,1,1,2,2,2),two=c(1,2,3,1,2,3), three=rnorm(6))
testdata

agg.results<-aggregate(testdata$three, list(testdata$one,testdata$two), mean)
tap.results<-tapply(testdata$three,list(testdata$one,testdata$two),mean)
class(agg.results)
class(tap.results)
```

### aggregate more
when the first argument of aggregate is a matrix or multicolumn dataframe, the statistics will be calculated for each column
```{r agg more}
testdata = cbind(sample(1:5,size=100,replace=TRUE),matrix(rnorm(500),nrow=100,ncol=5))
dimnames(testdata)=list(NULL, c("grp","A","B","C","D","E"))
aggregate(testdata[,-1],list(grp=testdata[,1]),min)
```

### split
```{r split}
data("Orange")
trees= split(Orange,Orange$Tree) # a list of dataframe, each has the same tree, and group by tree
sapply(trees,function(x)coef(lm(circumference~age,data=x)))
by(Orange,Orange$Tree,function(x) coef(lm(circumference~age,data=x)))
```

### if else
if (condition) expr
else other_expr

and ifelse
```{r ifelse}
z=c(7,10,-3,4,12,-5)
ifelse(z>0,z,0)
```

### paste
paste() can be used to convert its arguments to strings. When vectorized arguments are used, the collapse argument can concatenate the resulting vector to one single string. In addition, paste can convert numeric values to characters that can be used as the index for list/vectors.
```{r paste}
paste("x","equals 1","and y equals", "5",sep="")  #no vectorization arguments
paste("x",1:5, sep="") #vectorized arguments, obtain a vector
paste("x",1:5,sep="*", collapse="|") # conver the vector to a single string


x=c("10"=110,"7"=107,"12"=112)
is.vector(x)
x[paste(3+7)]
```

### substring
substring is to get the substrings from a string using the start and end indices by exact indexing. It can also replace the substring by reassigning substrings without changing the string length.
```{r substring}
animals=c("dog","cat","chicken","goat","sheep")
substring(animals,2,4)
substring(animals,2)

#replace the substring
pet="Rover"
substring(pet,2,3)="abcd" #ov were replaced by ab. cd was not used. substring will not change the string length
```

###strsplit
strsplit always returns a list. When a single string is input, it will return an one element list. Therefore, unlist() is usually used to convert the list to vector for vector operations
```{r strsplit}
strs=c("the razor toothed piranhas","of the genera Serrasalmus")
strsplit(strs," ") #use " " as the separator to return a list with 2 elements
strsplit(strs, c(" ","")) #use " " to split the first and "" for the second string

#if there is only one string, an one element list is returned
str1="the razor toothed piranhas"
strsplit(str1," ")
unlist(strsplit(str1," ")) # a vector is returned
```

#grep and regexpr
grep gives the index of the string element in a string vector that contains the pattern. To find the specific position of the pattern in each string, use regexpr. The vector returned by regexpr is the index vector of the pattern in strings.
```{r grep and regexpr}
states=c("Alabama","Alaska","California","michigan")
grep("a[lm]",states)
regexpr("a[lm]",states)
class(regexpr("a[lm]",states))
a<-regexpr("a[lm]",states)
```

###sub and gsub
sub and gsub replace the first argument (pattern as regular expression) in a string by the second argument. sub() only replace the first occurence, while the gsub() replace all of the occurences.
```{r sub gsub}
vars=c("date98","size98","x98weight98")
sub('98','04',vars)
gsub('98','04',vars)
```

### match and pmatch
match returns the index of each of the element in the first vector in the second vector. For elements in the first vector that do not appear in the second vector, the corresponding index is defined by nomatch argument, or NA.
```{r match pmatch}
match(1:10,c(7,2,5,4,3),nomatch=0)
1:10 %in% c(7,2,5,4,3) #for logical results

pmatch("m",c("mean","median","maximum"))
pmatch("^m",c("mean","median","maximum"))
pmatch("me",c("mean","median","maximum"))  #returns na since me is not unique
pmatch("ma",c("mean","median","maximum"))  #returns 3 since ma is unique to maximum
grep("m",c("mean","median","maximum"))     #compared to grep where all the indices of the matched pattern are returned
```

### merge matrices and dataframes
The default behavior of merge is inner join. By setting all=TRUE, all the rows including those not having the common join column values will be joined, with the NA value for the join column. In addition, the joined rows are ordered by the join column (set up using "by" argument). This automatic sorting can be turned off using sort=FALSE. When not sorting, the order is joined rows first, then the rows of the first input dataframe/matrix, then the second input.

When the join column names of the dataframes/matrices are different, use by.x= and by.y=. The all.x= and all.y= are used to include all rows in x or y. These all.x and all.y operations are same as left join and right join, and all.x and all.y take logical input (TRUE or FALSE). In addition, a vector of names or column numbers can be give for by=.
```{r merge}
one<-data.frame(a=c(1,7,9,3),b=c(12,18,24,19))
two<-data.frame(a=c(9,3,2,7,8),y=c(108,209,107,114,103))
merge(one,two,by="a")
merge(one,two,by="a",all=TRUE)
merge(one,two,by="a",all=TRUE,sort=FALSE)
merge(one,two,by="a",sort=FALSE)

three<-data.frame(d=c(1,7,9,3),b=c(12,18,24,19))
four<-data.frame(a=c(9,3,2,7,8),y=c(108,209,107,114,103))
merge(three,four, by.x="d",by.y="a")
merge(three,four, by.x="d",by.y="a",all.x=TRUE)

five<-data.frame(d=c(1,7,9,3),b=c(12,18,24,19),e=c(2,4,6,3))
six<-data.frame(a=c(9,3,2,7,8),y=c(108,209,107,114,103),z=c(6,3,7,9,4))
merge(five,six, by.x=c("d","e"),by.y=c("a","z")) #only rows with all the matched join columns will be included
```

### reshape
reshape() converts data between long and wide formats. The long data format has the following types of columns:
1. index columns, which will be used as the index column in wide format, and are defined by idvar=
2. time variables. Each unique combination of the time variable and index column corresponds to a row in the wide format. The rows are ordered by the time variable orders for each index column. The time variables are defined by timevar=
3. variables of interest. Each row corresponds to a variable of interst level will be expanded a column in the wide format. variable of interest is defined as v.names=
4. directions="long" or "wide"
5. varying= when converting to "long" format.
```{r reshape}
firmdata<-data.frame(firm=rep(c(7,12,19),times=c(3,3,3)),time=rep(c(1,2,3),3),x=c(7,19,12,13,18,9,21,15,7))
newdata=reshape(firmdata,idvar="firm",timevar="time",v.names="x",direction="wide")
longdata=reshape(newdata,direction="long")

widefirm=data.frame(firm=c(7,12,19),x1=c(7,13,21),x2=c(19,18,15),x3=c(12,9,7))
longfirm=reshape(widefirm,idvar="firm",varying=c("x1","x2","x3"),times=c(1,2,3),v.names="x",direction="long")
longfirm[order(longfirm$firm),]
```

### exapnd.grid
expand.grid() accepts vectors or lists of vectors, and create dataframe for each combination of the values from the vectors. The created dataframe can then be processed by apply() for grid serach or grid operations
```{r expand.grid}
gridx<-seq(-3,3,length=10)
gridy<-seq(-3,3,length=10)
gridx
gridy
values<-expand.grid(x=gridx,y=gridy)
head(values)
result=cbind(values,result=apply(values,1,function(z)sin(z[1])*cos(z[2])))
dim(result)
result[result[,3]==max(result[,3]),]
```

### probability distributions
q-quantile       input: probability; output: deviate, such as z value
p-probability    input:deviate;      output: probability
d-density        input:deviate;      output: density value
r-randoam number input:sample size;  output: vector of random numbers

norm:     Normal
pois:     Poisson
t:        Student's t
exp:      Exponential
binom:    Binomial
f:        F
gamma:    Gamma
chisq:    Chi-square
unif:     Uniform
```{r distributions}
qnorm(0.99,lower.tail=TRUE)   #1.28, 1.64, 1.96, 2.33 for 10, 5, 2.5, 1% probability all single sided
pnorm(1,lower.tail=TRUE)      #16%, 2.5%, 1% for 1, 2, 2.33 deviates all single sided
dnorm(1)                      #density value at deviate is 1
rnorm(10,mean=0,sd=1)         #generate 10 normal distributed numbers
```

### descriptive statistics
mean()
median()
range()
kurtosis()
skewness()
var()
mad()
sd()
IQR()
weighted.mean()
stem():   stem and leaf diagram
boxplot()
cor() use="all.obs","complete.obs","pairwise.complete.obs"
summary() provides minimum, mean, maximum,and quartiles
quantile() returns value of any desired quantiles

### Hypothesis tests
all the tests have alternative="two.sided","less" and "greater", and conf.level=
prop.test():      equality of proportions
binom.test():     one sample binomial
t.test():         students' t-test
cor.test():       correlation coefficient
wilcox.test():    Wilcoxon test
chisq.test():     Contingency tables
var.test():       Equality of variance
ks.test():        goodness of fit

### statistical models
lm():    linear model
glm:     generalized linear model
tree:    CART
nls:     non-linear models
aov:     analysis of variance
gam:     generalized additive models
cph:     cox proportional hazards
loess:   local regression models

functions displaying information about models are summary(), coef(), predict(), resid(), and plot(). These function should be implemented for all the models. 

common arguments to modeling functions:
formula=:    first argument
data=:       data frame to attach for model
subset=: logical experssion to select data to be used
weights=:    vector of weights to be applied to model
na.action=:  how to preprocess missing values
  -na.fail:  print error messages
  -na.omit:  default, omit missing value observations
  -na.pass:  do nothing
 
#### formulas in models
y~x1+x2+x3            additive model
y~x1+x2+x3-1          additive model without intercept
respons~.             response vs. all the other variables in the data frame
y~a*b*c               full factorial ANOVA model with a, b, c factors
y~x+I(x^2)+I(x^3)     polynomial regression


### three components of plots
1. high level functions that generate the entire plots including barplot(), boxplot(), contour(), dotchart(), hist(), pairs(), persp(), pie(), and plot()

2. low level functions that add to existing plots including abline(), arrows(), axis(), frame(),legend(),lines(), points(), rug(), symbols(), text(), and title()

3. graphics parameters used in either functions or the par function that change the layout or appearance of a plot. These parameters control things like margins, text size, tick marks, plotting style, and overall size of the plots.

### devices
the window of the monitor is the default device to display the plots. Other devices inculde postscript(), pdf(), bitmap(), and jpeg().

postscript(file="myplot.ps")
... plotting commands go here ...
dev.off()

dev.copy(device=jpeg, file="picture.jpg")
dev.off()

### multiple plots on a page
mfrow=c(nrow,ncol)
mfcol=c(nrow,ncol)
mfrow and mfcol will draw the plots in rows and columns, respectively. These functions can only be set in par() function. mai and mar set up the inner margin in inch and lines of text, respectively. oma and omi set up the outer margins of the plots in number of texts and inches, respectively. 
```{r multiple plots}
par(mfrow=c(2,2), oma=c(0.2,0.2,0.2,0.2), mai=c(0.2,0.2,0.2,0.2))
par(mfrow=c(2,2))
curve(sin(x),from=-pi, to=pi, main="Sine")
curve(cos(x),from=-pi, to=pi, main="Cosine")
curve(tan(x),from=-pi, to=pi, main="Tangent")
curve(sin(x)^2/cos(x),from=-pi, to=pi, main="Sin^2/Cosine")
```

### plot types and legends
polt() function accepts type= arguments, which can be "p"(points), "l"(lines), "b"(both), "s"(stairstep), and "n"(none). none will only show axes, allowing multiple lines to be plotted. 
legend() accepts x= and y= to specify the location of the legend and legend= to specify the text to appear in the legend. the title() is also used to add a title. 

mgp argument of the par() function is a vector of 3 values which specify the margin line for the axis title, axis labels and axis line.mpg(c(5,1,0))
```{r plot types and legends}
par(mfrow=c(1,1))
data(USArrests)
popgroup=cut(USArrests$UrbanPop,3,labels=c("Low","Medium","High"))
plot(range(USArrests$Murder),range(USArrests$Rape),type="n", xlab="Murder",ylab="Rape")
points(USArrests$Murder[popgroup=="Low"],USArrests$Rape[popgroup=="Low"],col="Red")
points(USArrests$Murder[popgroup=="Medium"],USArrests$Rape[popgroup=="Medium"],col="Green")
points(USArrests$Murder[popgroup=="High"],USArrests$Rape[popgroup=="High"],col="Blue")
legend(2,44,levels(popgroup),col=c("Red","Green","Blue"),pch=1)
title("Rape vs Murder by Population Density")
```

### Graphic parameters
Graphic parameters can be used in both par() and individual plotting routine. They can also be queried using par("parameter").
cex= character expansion (default=1)
pch= plotting character (using character or integer)
usr= vector of length 4 givering minimum and maximums of figure
col= colors-use names from the colors() function
srt= string rotation, in degrees colckwise
pin= vector with width and height of plotting region
fin= vector with width and height of figure region
lty= integer specifying line type


### plotting limits
using xlim=c(x, y), and ylim=c(x, y) to restrict plotting to a reduced range of points, or to expand the range of points.
```{r plotting limits}
data(airquality)
with(airquality,plot(Temp,Ozone))

# to display temperatures less than 80 and ozone less than 100, we could use
with(airquality,plot(Temp,Ozone,xlim=c(min(Temp,na.rm=TRUE),80),ylim=c(min(Ozone,na.rm=TRUE),100)))
```

### custom axes
axis() function allows creation of custom axes. xaxt= and yaxt= can be set at "n" to suppress the default creation of axes when producing a plot. side=(1=bottom, 2=left, 3=top, 4=right), at=, a vector of location for tick marks and labels, and labels= to specify the labels.

```{r axes}
data(USArrests)
popgroup=cut(USArrests$UrbanPop,3,labels=c("Low","Medium","High"))
rmeans=with(USArrests, aggregate(Rape,list(popgroup),mean))
where=barplot(rmeans[,2],xlab="Population Density") #where returns the center of the barplot
axis(1,at=where,labels=as.character(rmeans[,1])) #add the bottom axis
box()
title("Average Rape Rate by Population Density")
```

### the lattice library
The lattic library provides a complete set of functions for producing conditioning plots. In a conditioning plot, a data set is broken up into several parts, and each part is plotted on identical axes. The lattice library uses the formula notation of statistical models to describe the desired plot, adding the vertial bar to specify teh conditioning variable. All the scales of the subplots are identical, and the margins between the plots are eliminated to make it easy to compare.
```{r lattice}
data("Orange")
library("lattice")
xyplot(circumference~age|Tree,data=Orange,main="Size of Orange Trees",type="l")
```

### other lattice plots
In addition to xyplot(), there are following functions:
cloud()        3D scatter plot
qq()           Quantile-Quantile plots
dotplot()      Dot plots
wireframe()    3D Surface plots
histogram()    histogram
barchart()     bar charts
countourplot() contour plots
splom()        scatter plot matrices

Basic functions can also be used:
panel.abline()
panel.arrows()
panel.lines()
panel.points()
panel.segments()
panel.text()
```{r lattice complex panel}
library(lattice)
mypanel=function(x,y){
  panel.xyplot(x,y)
  panel.abline(lm(y~x))
}

mypanel1=function(x,y){
  panel.lines(x,y)
  panel.points(x,y)
}
xyplot(circumference~age|Tree,data=Orange,panel=mypanel1,main="Size of Orange Trees")
```

### Other options to Lattice functions
layout= a vector of 2 or 3 giving the number of rows, columns, and optionally, pages of the lattice display
as.table= TRUE, print panels in ascending order from left to right and top to bottom
subscripts= TRUE, prvides an argument called subscripts containing the subscripts of the observations
page= function to be called after each page of a multipage display. ??? Set to readline to imitate par(ask=TRUE)
